---
title:       "Data tidying"
author:      "Max Hachemeister"
date:        last-modified
date-format: iso
format:
  gfm:
    toc:     true
---

# Prerequisites

```{r}
#| label: setup
#| results: false
#| warning: false

library(tidyverse)
theme_set(theme_light())
```

# Introduction

## Tidy data

Rules of tidy data:

 1. Each variable is a column; each column is a variable.
 2. Each observation is a row; each row is an observation.
 3. Each value is a cell; each cell is a single value.
  
### Exercises
#### 1. For each of the sample tables, describe what each observation and each column represents.
    
 1. In `table1` each observation represents the population and tuberculosis cases per year per country, while the column represent the variables countries, year, cases and population.
    ```{r}
    table1
    ```
    
 2. In `table2` each observation represents either the number of cases or the population per year per country. Therefore the columns `type` represents two possible variables, namely cases or population.
    ```{r}
    table2
    ```
    
 3. In `table3` each observation represents the calculation of the tuberculosis rate per year per country. The columns `country` and `year` represent their literal variables, whereas `rate`, even though being a variable has two values in its cells.
    ```{r}
    table3
    ```
       
#### 2. Sketch out the process you'd use to calculate the `rate` for `table2` and `table3`.

> You will need to perform four operations:
>
> a. Extract the number of TB cases per country per year.
> b. Extract the matching population per country per year.
> c. Divide cases by population, and multiply by 1000.
> d. Store back in the approriate place.

 1. To calculate the `rate` from `table2` I would have to take the following steps:
    
    - Create separate columns for `cases` and `population`.
    - Move the values accordingly.
    - Mutate a `rate` column with "`cases` / `population`".
  
 2. To do get the `rate` from `table2` the following steps occur to me:
  
    - Mutate a `rate_result` column from the two values in the `rate` column.
      - Like "The number before '/' divided by the number after '/' ".
    - Basically it comes down to separating the two numbers before and after the `/`.
  
## Lengthening data

 - The most common task in tidying original data.
 - Used to convert many columns that are actually the same variable into two columns.
   - One being the common variable in which the old columns are now the values, and the other inheriting the values that were in the original columns.
 - So You are looking at four things in a sense.
 
 - Take note that within `pivot_longer()` the names for the *new* columns have to be quoted, aka. strings, as they do not yet refer to existing objects, like it would be case for `mutate()`.
 
 - Use `parse_number()` to extract the first number after a prefix string like `wk1`, `wk2`, `wk3`, etc. .
 
### How does pivoting work?

Create an example with `tribble()`: 
```{r}
df <- 
  tribble(
    ~id, ~bp1, ~bp2,
    "A", 100,  120,
    "B", 140,  115,
    "C", 120,  125
  )

df
```
 
Now pivot the tibble to have a column for the `number` of the blood pressure measurement and the `value` of that measurement.

```{r}
df |> 
  pivot_longer(
    cols = bp1:bp2,
    names_to = "measurement",
    values_to = "value"
  )
```

### Many variables in columns names

 - It is always sensible to describe all the different variables / variables that can be made out from looking at the raw data to get a feeling for how many columns there are and which name they might get.
 - In the `who2` example there are six columns / variables, including that describing the values that are given in the original columns.
 - Try to visualize how each original rows get repeatet by the number of original columns to be pivoted.
   - Simply because, and that's not so intuitive, no new data is created, but the data as is just rearranged.
   
### Data an variable names in the column headers

 - Okay, this is a complex case:
```{r}
household
```

 - The original columns describe several values of a single variable (`child`), and at the same time two other distinc variables (`name`, and `dob`).
 
 - `pivot_longer()` has special solution for that with the `.value` *sentinel* as a value for the `names_to` argument.
 
   - This way, each unique value of the position in the string you described with `.value` will become a new column which will be populated with the original values. Check it out:

```{r}
household |> 
  pivot_longer(
    # All columns, but `family`
    cols = !family,
    # Take the first value in the column string and separate unique names
    names_to = c(".value", "child"),
    # The column strings are separeted by "_"
    names_sep = "_",
    # Ignore rows that would have NA as value.
    values_drop_na = TRUE
  )
```

 - So now we see that the actual observation is not the family, but the single child.
 
## Widening data

 - For observations spread across multiple rows.
   - I.e. multiple 'variables' in a single column.
 - Less common, but more often in governmental contexts.
 
`pivot_wider()`s interface is the other way around, so you define the column in which those values are that should  be columns of their own.

The steps for `pivot_wider()` are as follows:

 - Identify the *unique* values in the defined column.
   - These become independent columns in a new dataframe which would be empty at first.
 - Then populate the new dataframe with the value of the original dataframe that belongs to the original row corresponding to the columns of the new dataframe.
 
Check this example:

Make a test dataframe
```{r}
df <- 
  tribble(
    ~id, ~measurement, ~value,
    "A", "bp1",        100,
    "B", "bp1",        140,
    "B", "bp2",        115,
    "A", "bp2",        120,
    "A", "bp3",        105
  )
```

Pivot this to a wider format so that the values of measurement become independent columns
```{r}
df |> 
  pivot_wider(
    names_from = measurement,
    values_from = value
  )
```

Check the unique values in the `measurement` column
```{r}
df |>
  distinct(measurement)
```

These become the new columns and for example the value of `blood` pressure for the first row would be put into the the new row:column combination with the matchting values for `id` and `measurement`.

```{r}
df |> 
  slice_head(n = 1)
```

## Summary

In this section I learned about the three rules for tidy data, which make total sense once understood, but are not always intuitive while collecting, or entering data. Therefore I also learned how to *read* untidy data, as well as the tools to transform this into the tidy dataframe I need for efficient exploration, namely `pivot_longer()` and `pivot_wider()`. I know now that if an actual single variable is spread out over multiple columns these would nee to be pivoted longer, while actually independent variables populating the same column need to be pivoted longer.
