---
title:       "Data import"
author:      "Max Hachemeister"
date:        last-modified
date-format: iso
format:
  gfm:
    toc:     true
---

# Prerequisites

```{r}
#| label: setup
#| results: false

library(tidyverse)
theme_set(theme_light())
```

## Reading data from a file

Get the `students` csv.
```{r}
students <- 
  read_csv("https://pos.it/r4ds-students-csv")

students
```

Have `"N/A"` be recognized as actual `NA`:
```{r}
students <- 
  read_csv(
    "https://pos.it/r4ds-students-csv",
    # Need to also add 'empty' fields as `NA`
    na = c("N/A", "")
  )

students
```
Convert row names from strings to values:
```{r}
students |> 
  rename(
    student_id = `Student ID`,
    full_name = `Full Name`
  )
```
Or use the `clean_names()` function from the `janitor` package which cleans all the other column names in the same call:
```{r}
students |> 
  janitor::clean_names()
```

`meal_plan` are a fixed set of known values, i. e. "factors". So convert the column type:
```{r}
students |> 
  janitor::clean_names() |> 
  mutate(meal_plan = factor(meal_plan))
```

`age` column is still character because one value is literal "five". Convert to numerical:
```{r}
students <- 
  students |> 
    janitor::clean_names() |> 
    mutate(
      meal_plan = factor(meal_plan),
      age = parse_number(if_else(age == "five", "5", age))
    )
```

### Other arguments

`read_csv` uses the first row as the column names by default:
```{r}
#| warning: false

read_csv(
  "a,b,c
  1,2,3
  4,5,6"
)
```

Sometimes there are metadata rows in the raw csv. Use the argument `skip = n` to drop the first `n` rows, or `comment = "#"` to drop lines with a given prefix:
```{r}
#| warning: false

read_csv(
  "Some first line
  And also a second line
  y,x,z
  1,2,3
  4,5,6",
  skip = 2
)

read_csv(
  ";; First line comment
  ;; Second line comment
  x,y,z
  1,2,3
  4,5,6
  7,8,9",
  comment = ";;"
)
```

The other way around use `col_names = FALSE` when there are no column names in the raw data:
```{r}
#| warning: false

read_csv(
  "1,2,3
  4,5,6
  7,8,9",
  col_names = FALSE
)
```
At this point you can also pass the column names as a character vector:
```{r}
read_csv(
  "1,2,3
  4,5,6
  7,8,9",
  col_names = c("x", "y", "z")
)
```

### Other file types

There are also special functions for other file types which are worth considering.
Check the  `read_csv` documentation.

### Exercises

#### 1. What function would you use to read a file where fields were separated with "|"?

The documentation for `read_csv()` states that the base function for this is `read_delim()` in which the delimiter can be specified:
```{r}
#| warning: false

read_delim(
  "x|y|z
  1|2|3
  4|5|6",
  delim = "|"
)
```

#### 2. Apart from `file`, `skip`, and `comment`, what other arguments do `read_csv()` and `read_tsv()` have in common?

Check the documentation again:
```{r}
?read_csv
?read_tsv
```

`read_csv()` and `read_tsv()` share all the same arguments, as they are variations of the same base function, that is `read_delim()`.

#### 3. What are the most important arguments to `read_fwf()`?

Check the documentation:
```{r}
?read_fwf
```

From what I've gathered `read_fwf()` need at least the filename or string, the `col_position` argument which defines where/how the columns are split up ,and the `col_names` argument wich can be part to the `col_position` values.

#### 4. Sometimes strings in a CSV file contain commas.

> To prevent them from causing problem, they need to be sorrounded by a quoting character, like `"` or `'`. By default, `read_csv()` assumes that the quoting character will be `"`. To read the following text into a data frame, what argument to `read_csv()` do you need to specify?
> `"x,y\n1,'a,b'"

I think I need to specify the non standard quote with the `quote` argument:
```{r}
#| warning: false

read_csv(
  "x,y\n1,'a,b'",
  quote = "\'"
)
```

#### 5. Identify what is wrong with each of the following incline CSV files. What happens when you run the code?

 1. Here we have a missing value for the third column which should be `c`.
    ```{r}
    #| warning: false

    read_csv("a,b\n1,2,3\n4,5,6")
    ```
    Interesting, so the values for the second and third column were concatenated into one instead of dropping the third column completely.

 2. This time there are more values in the third row than there are actual columns defined in the heading row. I guess as before the values will be concatenated so the second row will be `1,2,NA` and the third row will be `1,2,34`. Let's see:
    ```{r}
    #| warning: false

    read_csv("a,b,c\n1,2\n1,2,3,4")
    ```
   
 3. In this csv there is only one entry in the second row which is also half quoted with a `"` in front. I think this either won't work, or it will put a string `"1` as value: 
    ```{r}
    #| warning: false

    read_csv("a,b\n\"1")
    ```
    Yeah, it read the heading row, but then gave no further inputs. Odd, because I would think `"\"1"` would at least result in a string `"1`. 

 4. The rows of this csv have different data types; one is characters, and the other is integers. I think this will give a warning -or error-, or just keep the rows that are of the same type as the first one found.
    ```{r}
    #| warning: false
    
    read_csv("a,b\n1,2\na,b")
    ```
    Haha, sure it just reads all the values as strings. That's sensible.
    
 5. Here the separator is `;` instead of `,` which shouldn't bother the function. Ah well but as `read_csv()` is used it should not work.
    ```{r}
    #| warning: false

    read_csv("a;b\n1;3")
    ```
    Sure, just read is as strings and then single rows. :-)

So `read_csv()` will do anything to get the data into R. Probably so you can do the wrangling as soon as possible within R. Might be a good idea, but is something to be aware of.

#### 6. Practice referring to non-syntactic names in the following data frame by:

 a. Extracting the variable called `1`.
 b. Plotting a scatterplot of `1` vs. `2`.
 c. Creating a new column called `3`, which is `2` divided by `1`.
 d. Renaming the columns to `one`, `two`, and `three`.
 
```{r}
annoying <- 
  tibble(
    `1` = 1:10,
    `2` = `1` * 2 + rnorm(length(`1`))
  )
```

 a. Extracting the variable called `1`.
    ```{r}
    #| warning: false

    annoying |> 
      select(`1`)
    ```
  
 b. Plotting a scatterplot of `1` vs. `2`.
    ```{r}
    annoying |> 
      ggplot(
        aes(`1`, `2`)
      ) +
      geom_point()
    ```

 c. Creating a new column called `3`, which is `2` divided by `1`.
    ```{r}
    annoying |> 
      mutate(`3` = `2` / `1`)
    ```

 d. Renaming the columns to `one`, `two`, and `three`.
    ```{r}
    annoying |> 
      mutate(`3` = `2` / `1`) |> 
        rename(
          `one` = `1`,
          `two` = `2`,
          `three` = `3`
        )
    ```
    
## Controlling column types

When columns turn out to be character, even though you 'know' they are something different, specify to the column type and check the warning message to see why it was read as a string:

```{r}
simple_csv <-
  "x
  10
  .
  20
  30"

read_csv(
  simple_csv,
  col_types = "d"
)
```

Check the error:
```{r}
problems()
```

And correct it in the original call:
```{r}
read_csv(
  simple_csv,
  col_types = "d",
  na = "."
)
```

Yes, it also worked before, but adressing the warning messages helps you be aware of what you really want from the code.

## Reading data from multiple files

If the data is spread across several files, but at least with the same column layout, you can directly get them into a single dataframe by putting their paths into a vector:

```{r}
sales_files <- 
  c(
    "https://pos.it/r4ds-01-sales",
    "https://pos.it/r4ds-02-sales",
    "https://pos.it/r4ds-03-sales"
  )

read_csv(
  sales_files,
  id = "file"
)
```

To spare you typing all the filenames individually you can use this also with `list.files()` to get the list of all the files in a folder, that match what you are looking for. But filemanagement in R is a jam for its own.

## Writing to a file

Use `write_csv()` or `write_tsv()` to write your dataframes back to disk.
Problem is that these files do not save the column type. The `write_rds()` function solves that, but you can also use the `write_parquet()` function. Both save the data as binary files with metadata, which are faster to read.


## Data entry

Entry data column-wise:
```{r}
tibble(
  x = c(1, 2, 5),
  y = c("h", "m", "g"),
  z = c(.08, .83, .60)
)
```

Entry data row-wise:
```{r}
tribble(
  ~x, ~y, ~z,
  1, "h", .08,
  2, "m", .83,
  3, "g", .60
)
```

## Summary

In this chapter I've learned how to read data from `csv` files of different forms like comma-, tab- or something else-separated structures. I learned the different datatypes columns can have, how R idetifies those, and how I can troubleshoot if it did not work. I became aware of different file types to write my new dataframes back to, and their respective pros and cons. I now know how to read in files in batch, instead of typing them in manually.