---
title:       "Data import"
author:      "Max Hachemeister"
date:        last-modified
date-format: iso
format:
  gfm:
    toc:     true
---

# Prerequisites

```{r}
#| label: setup
#| results: false

library(tidyverse)
theme_set(theme_light())
```

## Reading data from a file

Get the `students` csv.
```{r}
students <- 
  read_csv("https://pos.it/r4ds-students-csv")

students
```

Have `"N/A"` be recognized as actual `NA`:
```{r}
students <- 
  read_csv(
    "https://pos.it/r4ds-students-csv",
    # Need to also add 'empty' fields as `NA`
    na = c("N/A", "")
  )

students
```
Convert row names from strings to values:
```{r}
students |> 
  rename(
    student_id = `Student ID`,
    full_name = `Full Name`
  )
```
Or use the `clean_names()` function from the `janitor` package which cleans all the other column names in the same call:
```{r}
students |> 
  janitor::clean_names()
```

`meal_plan` are a fixed set of known values, i. e. "factors". So convert the column type:
```{r}
students |> 
  janitor::clean_names() |> 
  mutate(meal_plan = factor(meal_plan))
```

`age` column is still character because one value is literal "five". Convert to numerical:
```{r}
students <- 
  students |> 
    janitor::clean_names() |> 
    mutate(
      meal_plan = factor(meal_plan),
      age = parse_number(if_else(age == "five", "5", age))
    )
```

### Other arguments

`read_csv` uses the first row as the column names by default:
```{r}
#| warning: false

read_csv(
  "a,b,c
  1,2,3
  4,5,6"
)
```

Sometimes there are metadata rows in the raw csv. Use the argument `skip = n` to drop the first `n` rows, or `comment = "#"` to drop lines with a given prefix:
```{r}
#| warning: false

read_csv(
  "Some first line
  And also a second line
  y,x,z
  1,2,3
  4,5,6",
  skip = 2
)

read_csv(
  ";; First line comment
  ;; Second line comment
  x,y,z
  1,2,3
  4,5,6
  7,8,9",
  comment = ";;"
)
```

The other way around use `col_names = FALSE` when there are no column names in the raw data:
```{r}
#| warning: false

read_csv(
  "1,2,3
  4,5,6
  7,8,9",
  col_names = FALSE
)
```
At this point you can also pass the column names as a character vector:
```{r}
read_csv(
  "1,2,3
  4,5,6
  7,8,9",
  col_names = c("x", "y", "z")
)
```

### Other file types

There are also special functions for other file types which are worth considering.
Check the  `read_csv` documentation.

### Exercises

#### 1. What function would you use to read a file where fields were separated with "|"?

The documentation for `read_csv()` states that the base function for this is `read_delim()` in which the delimiter can be specified:
```{r}
#| warning: false

read_delim(
  "x|y|z
  1|2|3
  4|5|6",
  delim = "|"
)
```

#### 2. Apart from `file`, `skip`, and `comment`, what other arguments do `read_csv()` and `read_tsv()` have in common?

Check the documentation again:
```{r}
?read_csv
?read_tsv
```

`read_csv()` and `read_tsv()` share all the same arguments, as they are variations of the same base function, that is `read_delim()`.

#### 3. What are the most important arguments to `read_fwf()`?

Check the documentation:
```{r}
?read_fwf
```

From what I've gathered `read_fwf()` need at least the filename or string, the `col_position` argument which defines where/how the columns are split up ,and the `col_names` argument wich can be part to the `col_position` values.

#### 4. Sometimes strings in a CSV file contain commas.

> To prevent them from causing problem, they need to be sorrounded by a quoting character, like `"` or `'`. By default, `read_csv()` assumes that the quoting character will be `"`. To read the following text into a data frame, what argument to `read_csv()` do you need to specify?
> `"x,y\n1,'a,b'"

I think I need to specifiy the non standard quote with the `quote` argument:
```{r}
#| warning: false

read_csv(
  "x,y\n1,'a,b'",
  quote = "\'"
)
```

