---
title:       "Layers"
author:      "Max Hachemeister"
date:        last-modified
date-format: iso
format:
  gfm:
    toc:     true
---

# Prerequisites

```{r}
#| label: setup
#| results: false

library(tidyverse)
theme_set(theme_light())
```

## Aesthetic mappings

### Exercises

#### 1. Create a scatterplot of `hwy` vs. `displ` where the points are pink filled in triangles.

I usually do `y` vs. `x` as a convention, where `y` is the variable to be explained and `x` the one in relation to that.
```{r}
mpg |> 
  ggplot(aes(x = displ, y = hwy)) +
  geom_point(
    shape = 17,
    color = "pink"
  )
```
Interestingly the "filled" triangle shape also uses the `color`, instead of `fill`.

#### 2. Why did the following code not result in a plot with blue points?

> ```{r}
> #| eval: false
> 
> ggplot(mpg) +
> geom_point(aes(x = displ, y = hwy, color = "blue))
> ```

This might be because the color is mapped as an aesthetic, hence dynamically changing instead of being set for every point. Let's see what happens when the `color` mapping is outside of the `aes()` function.
```{r}

mpg |> 
  ggplot(aes(x = displ, y = hwy), color = "blue")
```

Okay this did not work. The `aes()` function has to be in the actual `ggplot()` function, and not as part of the geom.

What happens, when the whole `aes()` call is put into the `ggplot()` function:
```{r}
mpg |> 
  ggplot(aes(x = displ, y = hwy, color = "blue")) +
  geom_point()
```

Almost there. I think the `aes()` always expects to be mapped to a column of a dataframe, or at least to a vector with several values. That's why just mapping the single value of `"blue"` to `color` does not result in actually blue points for the geom. 

I think I can either set the `"blue"` directly within the geom, or leave it in `ggplot()`, but outside of `aes()`, so that it will be passed down to all following geom calls.

I will try both here:

```{r}
# Specify color within geom
mpg |> 
  ggplot(aes(x = displ, y = hwy)) +
  geom_point(color = "blue")

# Set color in ggplot, but outside of aes() to be passed down.
mpg |> 
  ggplot(aes(x = displ, y = hwy), color = "blue") +
  geom_point()
```

Aha, It only works directly in the geom function. Good to know.

#### 3. What does the `stroke` aesthetic do? What shapes does it work with?
> Hint: use `?geom_point`

The documentation states that `stroke` defines the width of the "borders" of shapes that have these, like 21.

Let's see:
```{r}
# Make one version
mpg |> 
  ggplot(aes(displ, hwy)) +
  # Set is in the geom as I just learnedn
  geom_point(
    shape = 21,
    stroke = 1
  )

# And another version to see the difference
mpg |> 
  ggplot(aes(displ, hwy)) +
  # Set is in the geom as I just learnedn
  geom_point(
    shape = 21,
    stroke = 2
  )

```

#### 4. What happens if you map an aesthetic to something other than a variable name, like `aes(color = displ < 5)`?
> Note, you'll also need to specify x and y.

Let's see:
```{r}
mpg |> 
  ggplot(aes(displ, hwy, color = displ < 5)) +
  geom_point(
    # I kept this because I like the appearance
    shape = 21,
    stroke = 1
  )
```

Nice, just as I found in exercise 2, mapping an aesthetic in `aes()` works with at least two values. And the `displ < 5` check resulted in a vector with `TRUE` and `FALSE` values, which `aes()` in turn could interpret and map to the `color` aesthetic.

## Geometric objects

### Exercises

#### 1. What geom would you use to draw a line chart?
> A Boxplot? A histogram? An area chart?

You can find all the answers in the text of the sections above.

Linechart, `geom_line()`:
```{r}
mpg |> 
  ggplot(aes(displ, hwy)) +
  geom_line()
```

Yeah it's ugly, but it's a linechart nevertheless.

Boxplot, `geom_boxplot()`:
```{r}
mpg |> 
  # Using drv here, to one boxplot per drive train type.
  ggplot(aes(drv, hwy)) +
  geom_boxplot()
```

Histogram, `geom_histogram()`:
```{r}
mpg |> 
  ggplot(aes(hwy)) + 
  geom_histogram(binwidth = 2)
```

Okay, area chart ist new. Maybe something like `geom_area()`?
```{r}
mpg |> 
  # Added fill aesthetic to distinguish areas.
  ggplot(aes(hwy, displ, fill = drv)) +
  geom_area()
```

Also not pretty, but an area chart.

#### 2. Earlier in this chapter we used `show.legend` without explaining it:
> ```{r}
> #| eval: false
> 
> ggplot(mpg, aes(x = displ, y = hwy)) +
>   geom_smooth(aes(color = drv), show.legend = FALSE)
> ```

> What does `show.legend = FALSE` do here? What happens if you remove it? Why do you think we used it earlier?

What does the code do:
```{r}
#| message: false

# show.legend = FALSE
mpg |> 
  ggplot(aes(displ, hwy)) +
  geom_smooth(aes(color = drv), show.legend = FALSE)

# show.legend = TRUE
mpg |> 
  ggplot(aes(displ, hwy)) +
  geom_smooth(aes(color = drv), show.legend = TRUE)
```

The `show.legend` argument of the `geom_smooth()` function defines whether a legend for the lines is displayed. It was used earlier to drop the legend so that the three plots have the same general look.

#### 3. What does the `se` argument to `geom_smooth()` do?

The documentation `?geom_smooth()` states "Display confidence band..."

Let's see what it does:
```{r}
#| message: false

# se = TRUE
mpg |> 
  ggplot(aes(displ, hwy)) +
  geom_smooth(aes(color = drv), se = TRUE)

# se = FALSE
mpg |> 
  ggplot(aes(displ, hwy)) +
  geom_smooth(aes(color = drv), se = FALSE)
```

The `se` argument of the `geom_smooth()` defines whether the grey areas around the lines are shown.

#### 4. Recreate the R code necessary to generate the following graphs.
> Note that wherever a categorical variable is used in the plot, it's `drv`.

The top left plot is a scatterplot of black points with a smooth line without grey area on top:
```{r}
#| message: false

mpg |> 
  ggplot(aes(displ, hwy)) +
  geom_point() +
  geom_smooth(se = FALSE)
```

The top right plot is the same plot, this time with a smooth for each type of `drv`:
```{r}
#| message: false

mpg |> 
  ggplot(aes(displ, hwy)) +
  geom_point() +
  geom_smooth(aes(group = drv), se = FALSE)

```

The middle left plot is a scatterplot colored by `drv` with colored smooth lines for each `drv` on top as well, and as for this and all plots in this exercise the area around the smooth lines was dropped:
```{r}
#| message: false

mpg |> 
  ggplot(aes(displ, hwy, color = drv)) +
  geom_point() +
  geom_smooth(se = FALSE)
```

The middle right plot is the same as before but this time with only a single overall smooth line:
```{r}
#| message: false

mpg |> 
  ggplot(aes(displ, hwy)) +
  geom_point(aes(color = drv)) +
  geom_smooth(se = FALSE)
```

The bottom left plot is the same as before but now there are three smooth lines again, each with a different linetype:
```{r}
#| message: false

mpg |> 
  ggplot(aes(displ, hwy)) +
  geom_point(aes(color = drv)) +
  geom_smooth(aes(linetype = drv), se = FALSE)
```

And the final plot has no smooth lines but a white ring for each circle:
```{r}
mpg |> 
  ggplot(aes(displ, hwy)) +
  # Define it all especially for the shape.
  geom_point(
    shape = 21,
    aes(fill = drv),
    # Make it grey because the background is white.
    color = "grey90",
    # Make rings thicker
    stroke = 2)
```

